Redefining U--Pb discordance

### 2021-02-10 paper.tex
Incorporated reviewer comments.

\title{On the treatment of discordant detrital zircon U--Pb data}

@@

\begin{abstract}
  Detrital zircon U--Pb geochronology is a staple of sedimentary
  provenance analysis and crustal evolution studies. Constructing
  detrital age spectra is straightforward for concordant
  \textsuperscript{206}Pb/\textsuperscript{238}U- and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb-compositions.  But
  unfortunately, many detrital U--Pb datasets contain a significant
  proportion of discordant analyses. This paper investigates two
  decisions that must be made when analysing such discordant U--Pb
  data.

@@

  Applying these six discordance filters to a 10,000-grain dataset of
  detrital zircon U--Pb compositions reveals that: (i) the relative
  age discordance filter tends to suppress the young age components in
  U--Pb age spectra, whilst inflating the older age components; (ii)
  the Stacey-Kramers discordance filter is more likely to reject old
  grains and less likely to reject young ones; (iii) the p-value based
  discordance filter has the undesirable effect of biasing the results
  towards the least precise measurements; (iv) the logratio-based
  discordance filters are most strict for Proterozoic grains, and more
  lenient for Phanerozoic and Archaean age components; (v) of all the
  methods, the logratio distance to the concordia composition produces
  the best results, in the sense that it yields nearly identical
  \textsuperscript{206}Pb/\textsuperscript{238}U and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb age spectra without
  introducing an age bias.  All the methods presented in this paper
  have been implemented in the \texttt{IsoplotR} toolbox for
  geochronology.
\end{abstract}

@@

The U--Pb method consists of two paired decay systems, in which two
isotopes of the same radioactive parent (\textsuperscript{238}U and
\textsuperscript{235}U) decay to two isotopes of the same radiogenic
daughter (\textsuperscript{206}Pb and \textsuperscript{207}Pb,
respectively). This paired decay system provides a powerful internal
consistency check for the method, which is absent from other
chronometers. By `double dating' samples with the
\textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{235}U methods (or,
equivalently, the \textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb-methods) it is
possible to verify whether the isotopic system is free of primary or
secondary disturbances. The most reliable age constraints are obtained
from samples whose \textsuperscript{206}Pb/\textsuperscript{238}U,
\textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb ages are statistically
indistinguishable from each other. U--Pb compositions that fulfil this
requirements are `concordant'. Those that violate it are `discordant'.

Discordance can be caused by a number of mechanisms, including: (a)
the presence of non-radiogenic (`common') lead; (b) initial
disequilibrium between the short-lived nuclides of the
\textsuperscript{238}U--\textsuperscript{206}Pb and
\textsuperscript{235}U--\textsuperscript{207}Pb decay chains; (c)
partial loss of radiogenic lead during high grade metamorphism; and
(d) mixing of different age domains during micro-analysis
\citep{schoene2014}. These complicating effects can often be diagnosed
and remediated when multiple cogenetic crystals are available from the
same sample. If the aliquots form an isochron (or `discordia') line in
\textsuperscript{238}U-\textsuperscript{204,6,7}Pb space, then this
line can be used to recover robust chronologies from discordant data
\citep{ludwig1998}.

@@

Section~\ref{sec:discordance1} compares and contrasts existing
discordance filters based on age disparity and p-values. It shows that
the relative age definition strongly favours older samples over young
ones, and that the p-value definition, which has gained popularity in
recent years, hurts both the accuracy and precision of detrital
geochronology. The age disparity and p-value definitions are heuristic
by nature and are not based on firm statistical or geological
arguments.  Although they are essentially the only two definitions of
discordance that are used in detrital geochronology today, they are by
no means the only two possible options.

@@

Although the new definitions are arguably more attractive than the old
ones from a theoretical point of view, this does not guarantee that
they produce more sensible results. To test their performance on real
data, Section~\ref{sec:application} applies all six discordance
filters to a compilation of detrital zircon U--Pb data. Although the
true age distribution of this dataset is unknowable, the results
suggest that the logratio based discordance filters produce the most
accurate, and most easily interpretable results. The p-value
definition fares the worst and is not recommended.

@@

Both of these problems can be solved by using `hybrid' concordia ages
instead of `pure' \textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb ages. Concordia ages
are defined by \citet{ludwig1998} as the `most likely' (in a
statistical sense) U--Pb age given the isotopic ratio composition and
its analytical uncertainty (Figure~\ref{fig:concordia}).  Let $r_{75}$
and $r_{68}$ be the measured
\textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{206}Pb/\textsuperscript{238}U ratios, respectively,
and let $\sigma[r_{68}]^2$, $\sigma[r_{76}]^2$,
$\sigma[r_{68},r_{76}]$ be their (co)variances.  Then the concordia
age $t_c$ is obtained by numerically minimising the sum of squares
$S$:
\begin{equation}
  S = \left[
    \begin{array}{@{}c@{}}
      r_{75} - \exp(\lambda_{235}t_c) + 1 \\
      r_{68} - \exp(\lambda_{238}t_c) + 1
    \end{array}
    \right]^T\left[
    \begin{array}{@{}cc@{}}
      \sigma[r_{75}]^2 & \sigma[r_{75},r_{68}] \\
      \sigma[r_{75},r_{68}] & \sigma[r_{68}]^2
    \end{array}
    \right]^{-1}
  \left[
    \begin{array}{@{}c@{}}
      r_{75} - \exp(\lambda_{235}t_c) + 1 \\
      r_{68} - \exp(\lambda_{238}t_c) + 1      
    \end{array}
    \right]
  \label{eq:S}
\end{equation}

@@

\begin{figure}[t]
  \includegraphics[width=7.3cm]{TW-option-1234.pdf}
  \caption{Discordance cutoffs for four of the six discordance
    definitions discussed in Sections~\ref{sec:discordance1} and
    \ref{sec:discordance2}. The $d_{p}$ and $d_c$ criteria are not
    shown because they depend on the analytical uncertainty of the
    measurements, which may vary between studies. The top panel shows
    cutoffs of $d_r=20\%$ (relative age filter, dashed line),
    $d_t=300$~Ma (absolute age filter, dash-dot line), $d_{sk}=2\%$
    (Stacey--Kramers filter, dotted line) and $d_{a}=15\%$ (Aitchison
    distance, solid line) on a Tera-Wasserburg concordia diagram,
    which is plotted in logarithmic space to provide a more balanced
    view of the old and young ends of the time scale. The bottom panel
    presents the \textsuperscript{207}Pb/\textsuperscript{206}Pb --
    \textsuperscript{206}Pb/\textsuperscript{238}U age discordance
    along this line. This shows that the $d_r$-criterion favours young
    samples and the $d_{sk}$-criterion old samples, whereas the
    $d_a$-criterion is most strict for samples of intermediate (i.e.,
    Proterozoic) age. }
  \label{fig:agediscordance}
\end{figure}

The p-value definition affects grains differently depending on their
analytical precision \citep{nemchin2005}. For example, consider a
1.5~Ga zircon that is $d_r=1\%$ discordant. If this grain were
analysed by LA-ICP-MS with an analytical precision of 2\%, say, then
it would pass the chi-square test and be accepted as being
concordant. However, if that same grain were analysed by TIMS with a
precision of 0.2\%, then the p-value criterion would reject it as
being discordant (Figure~\ref{fig:TIMSvsLAICPMS}). It seems
fundamentally unfair that an imprecise analytical method would be
favoured over a precise one. This is a pertinent problem because
technical innovations are increasing the precision of all analytical
approaches to U--Pb geochronology.  As precision improves, so does the
ability to detect ever small degrees of discordance. Using the p-value
criterion, there may come a time when no detrital zircon passes this
filter.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{TW-sk.pdf}
    \caption{Using the \citet{stacey1975} common Pb model as a
      discordance criterion. This criterion assumes that the
      discordance is caused by linear mixing (hence, the linear scale
      of this Tera-Wasserburg plot) between radiogenic Pb
      (intersections of the mixing lines with concordia) and common Pb
      (intersection of the mixing lines with the vertical axis). The
      dashed line marks the 20\% ($= \delta_1/[\delta_1+\delta_2]$)
      discordance cutoff. This discordance filter, which must be
      applied \emph{before} making any actual common Pb correction, is
      more forgiving for young grains than it is for old grains. In
      this respect, it has the opposite effect of the relative age
      filter shown in Figure~\ref{fig:agediscordance}.  }
    \label{fig:SK}
\end{figure}

Section~\ref{sec:discordance1} reviewed three existing discordance
definitions.  This Section will introduce three new ones.  None of the
definitions discussed thus far encode any information about the
geological mechanisms behind the discordance. As explained in
Section~\ref{sec:intro}, common Pb is one of the most likely causes of
discordance. Using a mantle evolution model \citep[e.g.][]{stacey1975}
to approximate the isotopic composition of this common Pb, discordance
can be defined as:
\begin{equation}
  d_{sk} = 1 - \left[\frac{{}^{238}U}{{}^{206}Pb}\right]\bigg/\left[\frac{{}^{238}U}{{}^{206}Pb}\right]^\ast
  \label{eq:dsk}
\end{equation}

\noindent where $[{}^{238}U/{}^{206}Pb]^\ast$ is the
\textsuperscript{238}U/\textsuperscript{206}Pb-ratio of the
intersection between concordia and a straight line connecting the
\textsuperscript{238}U/\textsuperscript{206}Pb--\textsuperscript{207}Pb/\textsuperscript{206}Pb
measurement to the inferred mantle composition (Figure~\ref{fig:SK}).

The common Pb definition of discordance is more forgiving for young
grains than it is for old ones. Importantly, if the discordance is
caused by common Pb, then the
\textsuperscript{206}Pb/\textsuperscript{238}U,
\textsuperscript{207}Pb/\textsuperscript{206}Pb and concordia age
estimates are all negatively biased with respect to the true
age. However this bias can be removed by applying a common-Pb
correction \emph{after} the data have been filtered.

@@

To address this issue, it is useful to subject the U--Pb isotopic
ratio data to a logarithmic transformation. So instead of analysing
our data on a conventional Tera-Wasserburg concordia diagram, all
calculations can be done in
log(\textsuperscript{207}Pb/\textsuperscript{206}Pb) vs.
log(\textsuperscript{238}U/\textsuperscript{206}Pb) space. The
advantage of this transformation is that it produces values that are
free to range from $-\infty$ to $+\infty$. Within this infinite
dataspace, the Euclidean distance metric can be safely applied.

@@

There exists a vast body of statistical literature detailing the
theoretical and practical advantages of logratio analysis. A deeper
discussion of this topic falls outside the scope of this paper, but
the interested reader is referred to \citet{aitchison1986} and
\citet{pawlowsky2015} for further information. The Euclidean distance
between logratios is also known as the `Aitchison distance'.
Discordance can be redefined as the Aitchison distance from the
measured logratios to the concordia line. We here introduce two ways
to do so.  A first option is to simply measure the distance along a
perpendicular line to the concordia curve
(Figure~\ref{fig:aitchison}):
\begin{equation}
  d_{a} = dx(t_{68})
  \sin\!\left(\arctan\!\left[ \frac{dy(t_{76})}{dx(t_{68})} \right]
  \right)
  \label{eq:dperp}
\end{equation}

\noindent where
\begin{equation}
\begin{split}
  dx(t)  = & \log\!\left[\frac{{}^{206}Pb}{{}^{238}U}\right]-
  \log\!\left[\exp(\lambda_{238} t)-1\right] \\
  \mbox{~and~}
  dy(t) = & \log\!\left[\frac{{}^{207}Pb}{{}^{206}Pb}\right] -
      \log\!\left[\frac{{}^{235}U}{{}^{238}U}
        \frac{\exp(\lambda_{235}t)-1}{\exp(\lambda_{238}t)-1}
        \right]
  \end{split}
\end{equation}

This definition produces a parallel band around the concordia line in
logarithmic Tera-Wasserburg space. In contrast with $d_r$, $d_t$,
$d_{sk}$, the $d_a$ criterion is less strict at both the young and old
extremes of the geological timescale, and more strict during the
Proterozoic Era (Figure~\ref{fig:agediscordance}), when the U--Pb
method is most reliable.

@@

\section{Application to a compilation of detrital zircon U--Pb data}
\label{sec:application}

It is difficult to ascertain the mechanism causing discordance in any
particular zircon grain. Therefore, it is unclear which of the
definitions in Sections~\ref{sec:discordance1} and
\ref{sec:discordance2} is `correct'. All we can is do is apply the
methods to real samples and investigate their outcomes. This Section
will apply the six discordance filters to a compilation of 10,000
detrital zircon U--Pb analyses that were acquired at the London
Geochronology Centre (LGC) at University College London.

The data come from field areas on all seven continents. They were
acquired by LA-ICP-MS, on either an Agilent 7700x (early samples) or
an Agilent 7900 (recent samples) instrument, which was coupled to a
New Wave NWR193 excimer laser. Analytical conditions varied slightly
between different samples but generally used a 25 or 35 micron laser
spot, Ple\v{s}ovice zircon as a primary standard \citep{slama2008}
and, in most cases, GJ-1 zircon as a secondary standard
\citep{jackson2004}. Data reduction was done with GLITTER
\citep{griffin2008}. The data were not subjected to any common Pb
correction or other filters, apart from a visual inspection to remove
the most extreme outliers ($<1\%$ of the data).

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{KDE.pdf}
  \caption{Four superimposed kernel density estimates (with 60~Ma
    bandwidth) for 10,000 unfiltered detrital zircon U--Pb
    dates. The \textsuperscript{207}Pb/\textsuperscript{235}U,
    \textsuperscript{206}Pb/\textsuperscript{238}U and concordia age
    spectra are similar.  However the KDE of the
    \textsuperscript{207}Pb/\textsuperscript{206}Pb data stands
    apart from the other three curves. It deviates both at the young
    end of the age spectrum (which it suppresses), and at the old
    end (which it inflates).
  }
  \label{fig:KDE}
\end{figure}

Figure~\ref{fig:KDEs} applies the six discordance filters to this
database. In order to emphasise the difference between the six
discordance definitions whilst treating them on an equal footing, each
of the filters was adjusted until half of the data were removed. This
was achieved by discordance cutoffs of $d_t$ = 51~Myr, $d_r = 0.0618$,
$d_{p} = 0.084$, $d_{sk} = 0.00225$, $d_a = 0.0219$ and $d_c =
0.0297$.

There are noticeable differences between the density estimates.  As
expected from the theoretical considerations laid out in
Sections~\ref{sec:discordance1} and \ref{sec:discordance2}, the
relative age filter greatly suppresses the younger age components
($<1.5$~Ga) relative to the older parts of the age spectrum
($>1.5$~Ga). The \citet{stacey1975} filter has the opposite effect.
It suppresses the Archaean age component by $\sim$20\% whilst further
increasing the prominence of the Neoproterozoic and Phanerozoic modes.

The discordance definitions based on the absolute age difference and
logratio distances have a comparatively minor effect on the shape of
the age spectrum. All six discordance filters reduce the difference
between the \textsuperscript{207}Pb/\textsuperscript{206}Pb and
concordia age spectra except for the p-value filter, which actually
exacerbates the problem.

If the degree of similarity between the four chronometers is taken as
a measure of their success, then the concordia distance filter ($d_c$)
is the most successful discordance criterion. Apart from a minor
difference at the youngest end of the age spectrum, the four different
chronometers yield virtually identical age distributions after passing
through this filter.

\begin{figure*}[t]
  \includegraphics[width=12cm]{KDEs.pdf}
  \caption{Filtered U--Pb age spectra for the test data, removing the
    50\% most discordant grains according to the six discordance
    filters reviewed in this paper. Note how the concordia distance
    filter ($d_c$, fifth panel) exhibits the smallest difference
    between the four dating methods (including
    \textsuperscript{207}Pb/\textsuperscript{206}Pb), whereas the
    p-value filter ($d_p$, sixth panel) generates the greatest
    differences between them.  This observation suggests that the
    $d_c$ criterion is the best and the $d_p$ criterion the worst of
    the six methods. Also note how the relative age filter ($d_r$)
    strongly inflates the older half of the age spectrum at the
    expense of the younger half. The Stacey-Kramers filter ($d_{sk}$)
    has the opposite effect, as expected from
    Figure~\ref{fig:agediscordance}.}
  \label{fig:KDEs}
\end{figure*}

Figure~\ref{fig:KDEs} removed 50\% of the data, in order to emphasise
the differences between the six filters. In real applications, less
stringents discordance filters are usually applied. As mentioned in
the introduction, most current detrital zircon studies apply a
10\%--30\% relative age cutoff.  Using the test data, we can evaluate
the equivalent values for the $d_t$, $d_t$, $d_{sk}$, $d_a$ and $d_c$
criteria (Table~\ref{tab:conversiontable}). For example, a relative
age filter of 10\% removes the same fraction of the test data as an
absolute age filter with $d_t=83.5$~Ma, a Stacey-Kramers filter with
$d_{sk}=0.381$\%, a perpendicular Aitchison filter with $d_a=3.61$\%,
or a concordia distance filter with $d_c=3.67$\%.

The p-value discordance filter has been omitted from this comparison
for two reasons. First, the use of this filter is discouraged for
reasons that have been discussed before. Second, the p-value cutoffs
that are equivalent to any given relative age difference are highly
laboratory dependent, with precise equipment requiring different
$d_p$-cutoffs than imprecise instruments. The other five discordance
filters are more universally applicable. So using a different set of
test data should only make a relatively minor difference to the values
in Table~\ref{tab:conversiontable}.

>>>>

\title{Redefining U--Pb discordance}

@@

\begin{abstract}
  Zircon U--Pb geochronology is a staple of crustal evolution studies
  and sedimentary provenance analysis. Constructing detrital age
  spectra is straightforward for concordant
  \textsuperscript{206}Pb/\textsuperscript{238}U- and
  \textsuperscript{207}Pb/\textsuperscript{206}Pb-compositions.  But
  unfortunately, many detrital U--Pb datasets contain a significant
  proportion of discordant analyses. This paper investigates two
  decisions that must be made when analysing such discordant U--Pb
  data.

@@

  Applying these six discordance filters to a 70,869-grain dataset of
  zircon U--Pb compositions reveals that: (i) the relative age
  discordance filter tends to suppress the young age components in
  U--Pb age spectra, whilst inflating the older age components; (ii)
  the Stacey-Kramers discordance filter is more likely to reject old
  grains and less likely to reject young ones; (iii) the p-value based
  discordance filter has the undesirable effect of biasing the results
  towards the least precise measurements; (iv) the logratio-based
  discordance filters are most strict for Proterozoic grains, and more
  lenient for Phanerozoic and Archaean age components; (v) of all the
  methods, the logratio distance to the concordia composition produces
  the best results, in the sense that it produces age spectra that
  most closely match those of the unfiltered data: it sharpens age
  spectra but does not change their shape. Unfortunately, the relative
  age definition fares the worst according to this criterion.  All the
  methods presented in this paper have been implemented in the
  \texttt{IsoplotR} toolbox for geochronology.
\end{abstract}

@@

The U--Pb method consists of two paired decay systems, in which two
isotopes of the same radioactive parent (\textsuperscript{238}U and
\textsuperscript{235}U) decay to two isotopes of the same radiogenic
daughter (\textsuperscript{206}Pb and \textsuperscript{207}Pb,
respectively). This paired decay system provides a powerful internal
consistency check for the method, which is absent from other
chronometers. By `double dating' samples with the
\textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{235}U methods (or,
equivalently, the \textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb-methods) it is
possible to verify whether the isotopic system is free of primary or
secondary disturbances. The most reliable age constraints are obtained
from samples whose \textsuperscript{206}Pb/\textsuperscript{238}U,
\textsuperscript{207}Pb/\textsuperscript{235}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb ages are statistically
indistinguishable from each other. U--Pb compositions that fulfil this
requirement are `concordant'. Those that fail to meet it are
`discordant'.

Discordance can be caused by a number of mechanisms, including: (a)
the presence of non-radiogenic (`common') lead; (b) initial
disequilibrium between the short-lived nuclides of the
\textsuperscript{238}U--\textsuperscript{206}Pb and
\textsuperscript{235}U--\textsuperscript{207}Pb decay chains; (c)
partial loss of radiogenic lead during high grade metamorphism; and
(d) mixing of different age domains during micro-analysis
\citep{schoene2014}. These complicating effects can often be diagnosed
and remediated when multiple cogenetic crystals are available from the
same sample. If the aliquots form an isochron (or `discordia') line in
U--Pb isotope space, then this line can be used to recover robust
chronologies from discordant data \citep{ludwig1998}.

@@

Section~\ref{sec:discordance1} compares and contrasts existing
discordance filters based on age disparity and p-values. It shows that
the relative age definition strongly favours older samples over young
ones, and that the p-value definition, which has gained popularity in
recent years, hurts both the accuracy and precision of detrital
geochronology. The age disparity and p-value definitions are heuristic
by nature and are not based on firm statistical or geological
arguments.  Although they are the two most popular definitions of
discordance in use today, they are by no means the only two possible
options.

@@

Although the new definitions are arguably more attractive than the old
ones from a theoretical point of view, this does not guarantee that
they produce more sensible results. To test their performance on real
data, Section~\ref{sec:application} applies the discordance filters to
a compilation of detrital zircon U--Pb data. Although the true age
distribution of this dataset is unknowable, the results suggest that
the logratio based discordance filters produce the most accurate, and
most easily interpretable results. The relative age definition fares
the worst.

@@

Both of these problems can be solved by using `hybrid' concordia ages
instead of `pure' \textsuperscript{206}Pb/\textsuperscript{238}U and
\textsuperscript{207}Pb/\textsuperscript{206}Pb ages. Concordia ages
are defined by \citet{ludwig1998} as the `most likely' (in a
statistical sense) U--Pb age given the isotopic ratio composition and
its analytical uncertainty (Figure~\ref{fig:concordia}).  Let $r_{86}$
and $r_{76}$ be the measured
\textsuperscript{238}U/\textsuperscript{206}Pb and
\textsuperscript{207}Pb/\textsuperscript{206}Pb ratios, respectively;
let $\sigma[r_{86}]^2$, $\sigma[r_{76}]^2$ and $\sigma[r_{86},r_{76}]$
be their (co)variances; and let $R_{68}(t) = \exp(\lambda_{238}t) -
1$, $R_{75}(t) = \exp(\lambda_{235}t) - 1$, and $R_{58} =
{}^{235}\mbox{U/}^{238}\mbox{U}$.  Then the concordia age $t_c$ is
obtained by numerically minimising the sum of squares $S$\footnote{The
  same calculation can also be performed in Wetherill space, and is
  actually easier there.}:
\begin{equation}
  S = \left[
    \begin{array}{@{}c@{}}
      r_{86} - 1/R_{68}(t_c) \\
      r_{76} - R_{58} R_{75}(t_c)/R_{68}(t_c)
    \end{array}
    \right]^T\left[
    \begin{array}{@{}cc@{}}
      \sigma[r_{86}]^2 & \sigma[r_{86},r_{76}] \\
      \sigma[r_{86},r_{76}] & \sigma[r_{76}]^2
    \end{array}
    \right]^{-1}
  \left[
    \begin{array}{@{}c@{}}
      r_{86} - 1/R_{68}(t_c) \\
      r_{76} - R_{58} R_{75}(t_c)/R_{68}(t_c)
    \end{array}
    \right]
  \label{eq:S}
\end{equation}

@@

\begin{figure}[t]
  \includegraphics[width=7.3cm]{TW-option-1234.pdf}
  \caption{Discordance cutoffs for four of the six discordance
    definitions discussed in Sections~\ref{sec:discordance1} and
    \ref{sec:discordance2}. The $d_{p}$ and $d_c$ criteria are not
    shown because they depend on the analytical uncertainty of the
    measurements, which may vary between studies. The grey envelopes
    mark cutoff values of $d_r=20\%$ (relative age filter),
    $d_t=300$~Ma (absolute age filter), $d_{sk}=2\%$ (Stacey--Kramers
    filter) and $d_{a}=15\%$ (perpendicular Aitchison distance) on a
    Tera-Wasserburg concordia diagram, which is plotted in logarithmic
    space to provide a more balanced view of the old and young ends of
    the time scale. The $d_sk$ and $d_t$ envelopes are truncated where
    they cross over into physically impossible negative isotope ratio
    space.}
  \label{fig:agediscordance}
\end{figure}

The p-value definition affects grains differently depending on their
analytical precision \citep{nemchin2005}. For example, consider a
1.5~Ga zircon that is $d_r=1\%$ discordant. If this grain were
analysed by LA-ICP-MS with an analytical precision of 2\%, say, then
it would pass the chi-square test and be accepted as being
concordant. However, if that same grain were analysed by TIMS with a
precision of 0.2\%, then the p-value criterion would reject it as
being discordant (Figure~\ref{fig:TIMSvsLAICPMS}). It seems
fundamentally wrong that an imprecise analytical method would be
favoured over a precise one. This is a pertinent problem because
technical innovations are increasing the precision of all analytical
approaches to U--Pb geochronology.  As precision improves, so does the
ability to detect ever small degrees of discordance. Using the p-value
criterion, there may come a time when no detrital zircon passes this
filter.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{TW-sk.pdf}
    \caption{Using the \citet{stacey1975} common Pb model as a
      discordance criterion. This criterion assumes that the
      discordance is caused by linear mixing (hence, the linear scale
      of this Tera-Wasserburg plot) between radiogenic Pb
      (intersections of the mixing lines with concordia) and common Pb
      (intersection of the mixing lines with the vertical axis, see
      inset). The dashed line marks the 20\% ($=
      \delta_1/[\delta_1+\delta_2]$) discordance cutoff. This
      discordance filter, which must be applied \emph{before} making
      any actual common Pb correction, is more forgiving for young
      grains than it is for old grains. In this respect, it has the
      opposite effect of the relative age filter shown in
      Figure~\ref{fig:agediscordance}.  }
    \label{fig:SK}
\end{figure}

Section~\ref{sec:discordance1} reviewed three existing discordance
definitions.  This Section will introduce three new ones.  None of the
definitions discussed thus far encode any information about the
geological mechanisms behind the discordance. As explained in
Section~\ref{sec:intro}, common Pb is one of the most likely causes of
discordance. Using a mantle evolution model \citep[e.g.][]{stacey1975}
to approximate the isotopic composition of this common Pb, discordance
can be defined as:
\begin{equation}
  d_{sk} = 1 - r_{86}/r_{86}^\ast
  \label{eq:dsk}
\end{equation}

\noindent where $r_{86}^\ast$ is the
\textsuperscript{238}U/\textsuperscript{206}Pb-ratio of the
intersection between concordia and a straight line connecting the
\textsuperscript{238}U/\textsuperscript{206}Pb--\textsuperscript{207}Pb/\textsuperscript{206}Pb
measurement to the inferred mantle composition (Figure~\ref{fig:SK}).

The common Pb definition of discordance is more forgiving for young
grains than it is for old ones. Importantly, if the discordance is
caused by common Pb, then the
\textsuperscript{206}Pb/\textsuperscript{238}U,
\textsuperscript{207}Pb/\textsuperscript{206}Pb and concordia age
estimates are all positively biased with respect to the true
age. However this bias can be removed by applying a common-Pb
correction \emph{after} the data have been filtered.

Even though Equation~\ref{eq:dsk} is mathematically able to produce
negative discordance values, such values cease to have a geologically
meaningful interpretation, because it is impossible for minerals to
inherit negative amounts of common Pb. Thus it may be advisable to set
a minimum cutoff of $d_{sk}>0$ when using the Stacey-Kramers filter.

@@

To address this issue, it is useful to subject the U--Pb isotopic
ratio data to a logarithmic transformation. So instead of analysing
our data on a conventional Tera-Wasserburg concordia diagram, all
calculations can be done in
ln(\textsuperscript{207}Pb/\textsuperscript{206}Pb) vs.
ln(\textsuperscript{238}U/\textsuperscript{206}Pb) space. The
advantage of this transformation is that it produces values that are
free to range from $-\infty$ to $+\infty$. Within this infinite
dataspace, the Euclidean distance metric can be safely applied.

@@

There exists a vast body of statistical literature detailing the
theoretical and practical advantages of logratio analysis. A deeper
discussion of this topic falls outside the scope of this paper, but
the interested reader is referred to \citet{aitchison1986} and
\citet{pawlowsky2015} for further information. The Euclidean distance
between logratios is also known as the `Aitchison distance'.
Discordance can be redefined as the Aitchison distance from the
measured logratios to the concordia line. We introduce two ways to do
so here.  A first option is to simply measure the distance along a
perpendicular line to the concordia curve
(Figure~\ref{fig:aitchison}):
\begin{equation}
  d_{a} = dx(t_{68})
  \sin\!\left(\arctan\!\left[ \frac{dy(t_{76})}{dx(t_{68})} \right]
  \right)
  \label{eq:dperp}
\end{equation}

\noindent where
\begin{equation}
\begin{split}
  dx(t)  = & \ln\!\left[r_{86}\right] + \ln\!\left[R_{68}(t)\right] \\
  \mbox{~and~}
  dy(t) = & \ln\!\left[r_{76}\right] -
  \ln\!\left[R_{58}\frac{R_{75}(t)}{R_{68}(t)}\right]
  \end{split}
\end{equation}

This definition produces a parallel band around the concordia line in
logarithmic Tera-Wasserburg space. In contrast with $d_r$, $d_t$,
$d_{sk}$, the $d_a$ criterion is less strict at both the young and old
extremes of the geological timescale, and more strict during the
Proterozoic Eon, when the U--Pb method is most reliable.

@@

\section{Application to a compilation of zircon U--Pb data}
\label{sec:application}

It is difficult to ascertain the mechanism causing discordance in any
particular zircon grain. Therefore, it is unclear which of the
definitions in Sections~\ref{sec:discordance1} and
\ref{sec:discordance2} is `correct'. All we can do is apply the
methods to real samples and investigate their outcomes. This Section
will apply the discordance filters to a dataset of 70,869 zircon U--Pb
analyses that were acquired by Sensitive High Resolution Ion
Micro-Probe (SHRIMP) mass spectrometery and compiled by Dr. S.
Bodorkos of Geoscience Australia.

The dataset includes 1,665 sedimentary, igneous and metamorphic
samples, mostly from Australia but including some other locations as
well.  The data were acquired by a variety of instruments (including
SHRIMP-1, -2 and -RG) using a range of different reference materials,
and processed on a range of different data reduction software
(including Squid-1, -2 and Prawn/Lead). The data were not subjected to
any common Pb correction or other filters, and were saved in a
Tera-Wasserburg format with zero error correlation between the
\textsuperscript{238}U/\textsuperscript{206}Pb and
\textsuperscript{207}Pb/\textsuperscript{206}Pb ratio uncertainties.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{KDE.pdf}
  \caption{Four superimposed kernel density estimates (with 50~Myr
    bandwidth) for 70,869 unfiltered detrital zircon U--Pb dates. The
    \textsuperscript{207}Pb/\textsuperscript{235}U,
    \textsuperscript{206}Pb/\textsuperscript{238}U and concordia age
    spectra are similar.  However the KDE of the
    \textsuperscript{207}Pb/\textsuperscript{206}Pb data stands apart
    from the other three curves. It deviates both at the young end of
    the age spectrum (which it suppresses), and at the old end (which
    it inflates).  }
  \label{fig:KDE}
\end{figure}

Figure~\ref{fig:KDEs} applies five of the six discordance filters to
this database (the p-value filter was omitted for reasons given in
Section~\ref{sec:discordance1}). In order to emphasise the difference
between the discordance definitions whilst treating them on an equal
footing, each of the filters was adjusted until half of the data were
removed. This was achieved by discordance cutoffs of
$-18.6<d_t<46.0$~Myr, $-1.4<d_r<3.66$\%, $-0.11<d_{sk}<0.27$\%,
$-0.78<d_{a}<1.94$\%, and $-0.91<d_c<2.20$\%.

There are noticeable differences between the density estimates.  As
expected from the theoretical considerations laid out in
Sections~\ref{sec:discordance1} and \ref{sec:discordance2}, the
relative age filter greatly suppresses the younger age components
($<1.5$~Ga) relative to the older parts of the age spectrum
($>1.5$~Ga). The \citet{stacey1975} filter has the opposite effect.
It suppresses the Archaean age component by $\sim{50}$\% whilst
further increasing the prominence of the Neoproterozoic and
Phanerozoic modes.

The discordance definitions based on the absolute age difference and
logratio distances have a comparatively minor effect on the shape of
the age spectrum. The change in shape between the age spectrum of the
full (unfiltered) dataset and the age spectra of the filtered datasets
can be visually assessed on quantile-quantile plots and quantified
using the Kolmogorov-Smirnov (KS) statistic \citep{vermeesch2013}.

If the KS-misfit is taken as a measure of success, then the concordia
distance filter ($d_c$) is the most effective discordance
criterion. It `sharpens' the spectrum without changing the relative
prominence of the modes at 400, 1200, 1800, and 2500~Max.

\begin{figure*}[t]
  \includegraphics[width=12cm]{KDEs.pdf}
  \caption{Left: filtered U--Pb age spectra for the test data,
    removing the 50\% most discordant grains according to five
    discordance filters reviewed in this paper, shown as a kernel
    density estimate with 50~Myr bandwidth. The complete (unfiltered)
    dataset is shown in grey. Right: quantile-quantile plots comparing
    the filtered and unfiltered datasets. KS = the Kolmogorov-Smirnov
    statistic. The relative age filter ($d_r$) introduces the greatest
    and the concordia distance ($d_c$) the smallest bias,
    respectively.}
  \label{fig:KDEs}
\end{figure*}

Figure~\ref{fig:KDEs} removed 50\% of the data, in order to emphasise
the differences between the discordance filters. In real applications,
less stringent discordance filters are usually applied. As mentioned
in the introduction, most current detrital zircon studies apply a
10\%--30\% relative age cutoff.  Using the test data, we can evaluate
the equivalent values for the $d_t$, $d_{sk}$, $d_a$ and $d_c$
criteria (Table~\ref{tab:conversiontable}). For example, a relative
age filter of 10\% removes the same fraction of the test data as an
absolute age filter with $d_t=97$~Myr, a Stacey-Kramers filter with
$d_{sk}=0.62$\%, a perpendicular Aitchison filter with $d_a=4.1$\%, or
a concordia distance filter with $d_c=4.6$\%.

The p-value discordance filter has been omitted from this comparison
for two reasons. First, the use of this filter is discouraged for
reasons given in Section~\ref{sec:discordance1}. Second, the p-value
cutoffs that are equivalent to any given relative age difference are
highly laboratory dependent, with precise equipment requiring
different $d_p$-cutoffs than imprecise instruments. The other five
discordance filters are more universally applicable. So using a
different set of test data should only make a relatively minor
difference to the values in Table~\ref{tab:conversiontable}.
